---
title: "repeated n-gram detection"
author: "Anne Rutten"
date: "January 31, 2020"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## aim:
flag data points that are part of a sequence that is present at least twice in a given data set.

### usage:

either just copy the `findDuplicates` function and run it inside your script, or use the interactive bit below to load your own .csv file


### `findDuplicates()`: identify duplicated sequences in a dataset

* works on data vectors (e.g., a dataframe column in 'long' format)
* function generates n-grams for given vector, iteratively increasing the sequence length from the specified `min_length` until no more duplicate n-grams are present in the data
* data points that are part of an n-gram that is present in the data more than once are marked with an identifier specific to the sequence

#### please note:

* the function does not yet check for overlapping sequences within a specific n-gram length, i.e. a sequence "A A B A A B A" will count and mark n-gram "A A B A" as duplicated
* longer sequences will overwrite shorter sequences that they overlap with, i.e., in the above example,3-gram "A B A" will be overwritten by 4-gram "A A B A".


#### dependencies & constants
```{r, message=FALSE }
library(tidyverse)
library(ngram)
library(shiny)
library(DT)
library(colorspace)

# for test run: Raphaels filename
fn <- "/home/anne/Pardosa_mesocosm_activity_P_R.csv"
```

#### ducttape-and-tiewraps wrapper around ngram::ngrams()

* somevector: vector of values that needs to be tested for repeat sequences
* min_length: the minimum sequence length
* sep: in case this is a vector of strings, some separator that is not present in the data itself

```{r}

findDuplicates <- function(somevector, min_length, sep=" ") {
  # paste all values into one string for n-gram calculation
  values_as_string <- paste(somevector, collapse= sep)
  charactervector <- as.character(somevector)
  output <- data.frame(value=somevector, seqID = NA)
  done <- FALSE
  i <- min_length
  while (!done) {
    duplicates <- get.phrasetable(ngram(values_as_string, i)) %>% 
                   filter(freq>1) %>%
                   select(ngrams) %>%
                   mutate(ngrams = strsplit(ngrams, sep))
    if (nrow(duplicates) > 0) {
      for(k in 1:nrow(duplicates)) {
        ng <- duplicates$ngrams[[k]]
        idx <- which(charactervector == ng[1])
        seqStart <- idx[sapply(idx, function(j) all(charactervector[j:(j+(length(ng)-1))] == ng))]
        output$seqID[rep(seqStart, each = i) + 0:(i-1)] <-paste0("n",i,"Seq",k)
     }
    i<-i+1
    } else done=TRUE
  }
  output$seqID
}
 
```

## the interactive bit:

1. load a csv
2. identify the column range for which you want to flag duplicate n-grams (these will be transformed to `long` format)
3. check the output (use the filter function to find specific sequences)

the sequence id is generated as follows: <n>Seq<ngramID>, where `n` is the sequence length, and `ngramID` does not carry other information than that they're separate entities.

##### example usage in Raphaels data:

download from Raphaels github: https://github.com/rroyaute/Royaute-Pruitt-Ecology-2015-Data-and-Code/blob/master/Data%20files/Pardosa_mesocosm_activity_P_R.csv

```{r, echo=FALSE}
shinyApp(
  ui = fluidPage(
      h3("1: load datafile"),
    inputPanel(
        column(12,textAreaInput("filename", label ="filename: ",value=fn, width = 500
                ))),
     inputPanel(actionButton("readRaw", label = "get data")),
    
  
h3("2: select column range of data to detect sequences for"),
inputPanel(     numericInput("fromCol", label = "focal column range start: ", value= 3),
                      numericInput("toCol", label = "focal column range end: ", value=10),
                      numericInput("minGram", label = "minimum sequence length ", value=4),
                      actionButton("run", label = "detect duplicates")
    ),
    h3("raw data:"),
    DT::dataTableOutput("rawd"),
    h3("with duplicated sequence IDs:"), 
    DT::dataTableOutput("dataSummary"),
    DT::dataTableOutput("duplicateNgramsWide")
    
  ),

server = function(input, output, session) {
  # load data 
  loadData <- eventReactive(input$readRaw,{
    dataLoaded <- TRUE
    read.csv(input$filename)
    })
  
  output$rawd <- DT::renderDataTable({loadData()})
  
  # flag duplicate n-grams
  
  duplicateNgrams <- eventReactive(input$run, {
    d <-loadData()
    longd <- gather(d, key, value, input$fromCol:input$toCol)
    longd$ngramID <- findDuplicates(longd$value, input$minGram)
    longd
    })
  
  output$duplicateNgramsWide <- DT::renderDataTable({
    duplicateNgrams <- duplicateNgrams()
     ids <- unique(duplicateNgrams$ngramID[!is.na(duplicateNgrams$ngramID)])
    cols <- rainbow_hcl(length(ids))
    duplicateNgramsWide <- duplicateNgrams%>% pivot_wider(names_from =key, values_from=c(value, ngramID)) 
    varnames <- names(duplicateNgramsWide)
    varsToFormat <- varnames[grepl("value", varnames)]
    varsRef <- varnames[grepl("ngramID", varnames)]
    
    datatable(duplicateNgramsWide) %>%
      formatStyle(varsToFormat, varsRef, backgroundColor=styleEqual(ids,cols))
  })
  output$duplicateNgrams <- DT::renderDataTable({
   duplicateNgrams <- duplicateNgrams()
     ids <- unique(duplicateNgrams$ngramID[!is.na(duplicateNgrams$ngramID)])
    cols <- rainbow_hcl(length(ids))
    datatable(duplicateNgrams) %>%
    formatStyle("ngramID", backgroundColor = styleEqual(ids,cols))
  })
  
  #summary
  dataSummary <-reactive({
     duplicateNgrams() %>%
      summarise(n_duplicates = sum(!is.na(ngramID)),
                n_total = n(),
                fraction = round(n_duplicates/n_total,2))
  })
  output$dataSummary <- DT::renderDataTable({
   dataSummary()
  })
},
options = list(height=5000))
  
```

