---
title: "static sequenceSniffer"
author: "Anne Rutten"
date: "February 1, 2020"
output: html_document
---

static version of the duplicated-n-gram-flagging example. Easier to copy functionality out of.

## aim:
flag data points that are part of a sequence that is present at least twice in a given data set.

### `findDuplicates()`: identify duplicated sequences in a dataset

* works on data vectors (e.g., a dataframe column in 'long' format)
* function generates n-grams for given vector, iteratively increasing the sequence length from the specified `min_length` until no more duplicate n-grams are present in the data
* data points that are part of an n-gram that is present in the data more than once are marked with an identifier specific to the sequence

#### please note:

* the function does not yet check for overlapping sequences within a specific n-gram length, i.e. a sequence "A A B A A B A" will count and mark n-gram "A A B A" as duplicated
* longer sequences will overwrite shorter sequences that they overlap with, i.e., in the above example,3-gram "A B A" will be overwritten by 4-gram "A A B A".

### usage:

*either just copy the `findDuplicates` function and run it inside your script, or

1. change the constants below and 
2. knit this document to generate a report for a specific datafile

#### dependencies

```{r, message=FALSE }
library(tidyverse)
library(ngram)
library(shiny)
library(DT)
library(colorspace)
library(ggplot2)
```


#### controls

```{r, message=FALSE }

# path to csv datafile
fn <- "/home/anne/Pardosa_mesocosm_activity_P_R.csv"

# column range of the focal data

fromColumn <- 3
toColumn <- 10

# minimum sequence length to consider

minLength <- 4

# if you want to reorder datapoints within some group and analyze the distribution of 'number of datapoints in duplicated sequences' in your dataset, specify grouping variable here
# the variable name that is assigned to the `long` concatenated columns specified above is 'value' 
groupingVars <-c("Treatment", "value")
```

#### ducttape-and-tiewraps wrapper around ngram::ngrams()

* somevector: vector of values that needs to be tested for repeat sequences
* min_length: the minimum sequence length
* sep: in case this is a vector of strings, some separator that is not present in the data itself

```{r}

findDuplicates <- function(somevector, min_length, sep=" ") {
  # paste all values into one string for n-gram calculation
  values_as_string <- paste(somevector, collapse= sep)
  charactervector <- as.character(somevector)
  output <- data.frame(value=somevector, seqID = NA)
  done <- FALSE
  i <- min_length
  while (!done) {
    duplicates <- get.phrasetable(ngram(values_as_string, i)) %>% 
                   filter(freq>1) %>%
                   select(ngrams) %>%
                   mutate(ngrams = strsplit(ngrams, sep))
    if (nrow(duplicates) > 0) {
      for(k in 1:nrow(duplicates)) {
        ng <- duplicates$ngrams[[k]]
        idx <- which(charactervector == ng[1])
        seqStart <- idx[sapply(idx, function(j) all(charactervector[j:(j+(length(ng)-1))] == ng))]
        output$seqID[rep(seqStart, each = i) + 0:(i-1)] <-paste0("n",i,"Seq",k)
     }
    i<-i+1
    } else done=TRUE
  }
  output$seqID
}
 
```

#### raw data: 

```{r, echo=FALSE}
# read data
d <- read.csv(fn)

# reformat d from wide to long
longd <- gather(d, key, value, fromColumn:toColumn)

#identify duplicate ngrams
longd$ngramID <- findDuplicates(longd$value, minLength)

# format back to wide
duplicateNgramsWide <- longd %>% pivot_wider(names_from =key, values_from=c(value, ngramID)) 

datatable(d)
```


#### data summary: 

```{r, echo=FALSE}

  #summary
  dataSummary <-longd %>%      
                summarise(n_duplicates = sum(!is.na(ngramID)),
                n_total = n(),
                fraction = round(n_duplicates/n_total,2))
dataSummary
```

#### detected sequences:

```{r, echo=FALSE}

#make a databale with highlighted background colour for duplicates

ids <- unique(longd$ngramID[!is.na(longd$ngramID)])
cols <- rainbow_hcl(length(ids))
varnames <- names(duplicateNgramsWide)
varsToFormat <- varnames[grepl("value", varnames)]
varsRef <- varnames[grepl("ngramID", varnames)]
    
markedTable <- datatable(duplicateNgramsWide) %>%
      formatStyle(varsToFormat, varsRef, backgroundColor=styleEqual(ids,cols))

markedTable
```

#### expected datapoints-in-duplicated-sequences distribution for this dataset:
* randomly reorder datapoints within the grouping level specified in `groupingVars`


```{r, echo=FALSE}

#resample within treatment

groupedD <- group_by_at(longd, groupingVars)

res <- list()
for (i in 1:1000) {
  newd <- groupedD %>%
    mutate(newOrder = runif(n())) %>%
    arrange(newOrder)
  
  newDupes <- findDuplicates(newd$value, minLength)
  res[[i]] <- sum(!is.na(newDupes))
}

res <- do.call(rbind, res)

summary(res)
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
resdf <- bind_rows(data.frame(what = "simulated", n_duplicated = res),
                   data.frame(what = "actual", n_duplicated = dataSummary$n_duplicates)
                   )

```

```{r, echo=FALSE}

ggplot(as.data.frame(resdf), aes(n_duplicated, colour = what, fill=what)) +
  geom_bar() +
  theme_bw() +
  labs(title = "number of datapoints that are part of a duplicate sequence",
       subtitle = paste("data reordered within:", paste(groupingVars, collapse=", "), "; n_runs=1000; minimum n-gram length =", minLength))
```